Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures

### 2020-09-16 manuscript/manuscript.Rmd
submission to jrn

---
title: Mokken scale analysis of the perioperative thirst discomfort scale
author:
  - name: ""
date_subtitle: "Blinded for peer review"
lead_author_surname: Blinded for peer review
doi_footer: "Pre-print submitted to *Applied Nursing Research*"
abstract: |
  *Introduction:* Thirst discomfort is common because patients are required to undergo long periods of fasting before medical and surgical procedures. The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale (PTDS) for measuring thirst discomfort before procedures.
  *Methods:* Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study. Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS. PTDS scores were compared with fasting duration to evaluate construct validity. Convergent validity was evaluated by comparing the PTDS with global thirst discomfort and thirst intensity ratings.
  *Results:* Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering. Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of ‘thick’ saliva and a ‘thick’ tongue. Scale reliability was adequate (rho = 0.84). There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (tau = 0.54; 95% CI = 0.43 to 0.63), as well as thirst intensity (tau = 0.49; 95% CI = 0.38 to 0.59). Duration of fasting was not associated with PTDS-5 scores.
  *Conclusion:* The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
keywords:
  - fasting
  - anesthesia
  - thirst discomfort
  - patient-reported outcomes
  - scale validation
  - mokken scale analysis
bibliography: references.bib
footer_contents: "Perioperative thirst discomfort scale"
#one_column: true
fontsize: 12pt
output: pinp::pinp
params: 
  word: false
  pdf: true
header-includes: >
  \usepackage[switch]{lineno}
  \linenumbers
  \usepackage{setspace}
  \onehalfspacing

@@

We did not observe an association between PTDS-5 and fasting duration for either food or fluids. This would suggest that it is important to assess thirst discomfort periodically, regardless of the duration of fasting. If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented. There is a lack of evidence regarding the efficacy of interventions to alleviate perioperative thirst discomfort. However, recent randomized controlled trials have identified clinically relevant redictions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].

>>>>

---
title: Validating the perioperative thirst discomfort scale for measuring thirst discomfort prior to procedures
author:
  - name: ""
date_subtitle: "Blinded for peer review"
lead_author_surname: Blinded for peer review
doi_footer: "Pre-print submitted to *Applied Nursing Research*"
abstract: |
  *Introduction:* Thirst discomfort is common because patients are required to undergo long periods of fasting before medical and surgical procedures. The aim of this study was to examine the validity and reliability of the perioperative thirst discomfort scale (PTDS) for measuring thirst discomfort before procedures.
  *Methods:* Fasting patients who were scheduled for an elective cardiac or interventional radiology procedure were included in a prospective observational study. Mokken scaling analysis was used to investigate the dimensionality and hierarchical nature of the PTDS. PTDS scores were compared with fasting duration to evaluate construct validity. Convergent validity was evaluated by comparing the PTDS with global thirst discomfort and thirst intensity ratings.
  *Results:* Five items from the perioperative thirst discomfort scale (PTDS-5) formed a Mokken scale with evidence of invariant item ordering. Participants most easily endorsed the item related to the desire to drink water through to first endorsing symptoms related to dryness of the mouth and lips before those related to the abnormal sensations of ‘thick’ saliva and a ‘thick’ tongue. Scale reliability was adequate (rho = 0.84). There was a positive correlation between PTDS-5 scores and the global thirst discomfort rating (tau = 0.54; 95% CI = 0.43 to 0.63), as well as thirst intensity (tau = 0.49; 95% CI = 0.38 to 0.59). Duration of fasting was not associated with PTDS-5 scores.
  *Conclusion:* The items in the PTDS-5 form a strong Mokken scale, meaning it is a reliable and precise way to order patients according to their thirst discomfort.
keywords:
  - fasting
  - anesthesia
  - thirst discomfort
  - patient-reported outcomes
  - scale validation
  - mokken scale analysis
bibliography: references.bib
footer_contents: "Perioperative thirst discomfort scale"
#one_column: true
fontsize: 12pt
output: pinp::pinp
params: 
  word: false
  pdf: true
header-includes: >
  \usepackage[switch]{lineno}
  \linenumbers
  \usepackage{setspace}
  \onehalfspacing

@@

We did not observe an association between PTDS-5 and fasting duration for either food or fluids. This would suggest that it is important to assess thirst discomfort periodically, regardless of the duration of fasting. If thirst discomfort is present, and fasting is still indicated, interventions to alleviate symptoms should be implemented. There is a lack of evidence regarding the efficacy of interventions to alleviate perioperative thirst discomfort. However, recent randomized controlled trials have identified clinically relevant reductions in thirst discomfort with menthol chewing gum and 30mL menthol popsicles [@garcia2019; @nascimento2020].
